Stream Members Only:\\ Data-Driven Characterization of Stellar Streams with Mixture Density Networks

### 2023-07-11 src/tex/ms.tex
[showyourwork] paper
Signed-off-by: nstarman <nstarman@users.noreply.github.com>

\documentclass[twocolumn]{aastex631}
\usepackage{showyourwork}
\usepackage{amsmath}
\usepackage{xcolor}
\usepackage{booktabs}
\usepackage{todonotes}

\newcommand{\mrm}[1]{\mathrm{#1}}
\newcommand{\mbs}[1]{\boldsymbol{#1}}
\newcommand{\mbf}[1]{\mathbf{#1}}
\newcommand{\mbb}[1]{\mathbb{#1}}
\newcommand{\mfk}[1]{\mathfrak{#1}}
\newcommand{\mcal}[1]{\mathcal{#1}}
\newcommand{\unit}[1]{[\text{#1}]}

\newcommand{\nth}[1]{{#1}_{\mrm{n}}}
\newcommand{\fth}[1]{{#1}_{\mrm{f}}}
\newcommand{\qth}[1]{{#1}_{\mrm{q}}}

\newcommand{\pdf}{P}
\newcommand{\cdf}{\Phi}

\newcommand{\Exp}[1]{e^{#1}}
\newcommand{\Erf}[1]{\mathrm{erf}\left(#1\right)}

\newcommand{\sigobs}{{\sigma_*}}
\newcommand{\JN}[1]{{\textcolor{red}{JN: #1}}}

\NewDocumentCommand \dif {o m} {
	\IfNoValueTF{#1}
		{ \mathrm{d}{#2} }
		{ \mathrm{d}^{#1}{\!#2} }
}

@@

% Title
\title{Characterizing Stellar Streams by ML}

% Author list
\author{Nathaniel Starkman}
\author{Jacob Nibauer}
\author{Jo Bovy}
\author{Jeremy Webb}

% Abstract with filler text
\begin{abstract}
    Stellar stream membership likelihoods with ML. It works!
\end{abstract}

@@

    [Content Here]

    \

    Paper Map:
    \begin{enumerate}
        \item Intro
        \item Method \begin{enumerate}
            \item Fig 1: Overall explainer figure that illustrates the method with Neural networks, and norm flow for CMD
            \item Fig 2: Bayesian graphical model
            \item Test On Synthetic Stream:
            \item Some sort of plot demonstrating successful recovery
        \end{enumerate}
        \item Application / Results \begin{enumerate}
            \item Test case GD-1: \begin{enumerate}
                \item CMD norm flow background
                \item Fig: Stream phi1 phi2 top panel and histograms of each dimension below GD-1
                \item Some plot in phi1-phi2 that shows density variations
            \end{enumerate}
            \item Test case Pal-5 \begin{enumerate}
                \item CMD norm flow background
                \item Fig: Stream phi1 phi2 top panel and histograms of each dimension below Pal 5
                \item Some plot in phi1-phi2 that shows density variations
            \end{enumerate}
        \end{enumerate}
        \item Conclusion
        \item Appendix
    \end{enumerate}

@@

\section{Method} \label{sec:method}

    Stellar streams

    \subsection{Likelihood Setup}\label{sub:likelihood_setup}

        \begin{figure}
        \centering
        \includegraphics[scale=0.5]{figures-not-syw/DAG_StreamDensity.pdf}
        \caption{Fill in.}
        \label{fig: DAG}
        \end{figure}

        We model the stream and background as functions continuous in some affine parameter. In principle this parameter can be anything. Though it might seem the best parameter is the stream's arc length, the most common, easiest, and practical parameter is the longitude -- $w_{\cdot, j\equiv0} \equiv \phi_1$ -- in some stream-oriented reference frame \citep[for details see][section 2]{2023MNRAS.522.5022S}.

        We model the likelihood for a single star as
        \begin{align} \label{eq:general_model}
            \pdf(\nth{\mbs{w}}, \nth{\mbs{m}} | \mbs{\theta})
                &= \sum_{q} \ f_q(w_{i, 0}) \pdf(\nth{\mbs{w}}|\mbs{\theta},q_i\!=\!q) \pdf(\nth{\mbs{m}}|\mbs{\theta},q_i\!=\!q)
        \end{align}
        where we have assumed that $\nth{\mbs{w}}$ and $\nth{\mbs{m}}$ are conditionally independent, and that the mixture weights $f$ are normalized s.t. $\sum_q f(w_{i,0}) = 1$. We discuss the mixture weights further in \autoref{sub:mixture_weight}

        Assuming the measurement of each star is independent, the total log-likelihood is a sum over all stars:
        \begin{equation}
            \ln\mcal{L}\left(\{(\nth{\mbs{w}},\nth{\mbs{m}})\} | \mbs{\theta}\right) = \sum_i  \ln \pdf(\nth{\mbs{w}}, \nth{\mbs{m}} | \mbs{\theta}).
        \end{equation}

        It will prove convenient to distinguish the models for $q=0,1$ as the \textit{background} and \textit{stream} models, respectively. In this notation:

        \begin{multline}
            \pdf(\nth{\mbs{w}}, \nth{\mbs{m}} | \mbs{\theta}) 
            = \\\phantom{+} \ \quad f \quad\ P^{(S)}(\nth{\mbs{w}}|\mbs{\theta}) P^{(S)}(\nth{\mbs{m}}|\mbs{\theta})  \\
            \quad + (1-f) P^{(B)}(\nth{\mbs{w}}|\mbs{\theta}) P^{(B)}(\nth{\mbs{m}}|\mbs{\theta}), \nonumber
        \end{multline}

        The probablity of a being a stream member is then simply
        \begin{equation}\label{eq:membership_prob}
            \pdf\left(S | \mbs{w}_n, \mbs{m}_n, \mbs{\theta} \right) = \frac{f P^{(S)}(\nth{\mbs{w}}|\mbs{\theta}) P^{(S)}(\nth{\mbs{m}}|\mbs{\theta}) }{ \pdf(\nth{\mbs{w}}, \nth{\mbs{m}} | \mbs{\theta})}.
        \end{equation}
    Hereafter, when we refer to ``membership probability," Eq.~\ref{eq:membership_prob} provides the formal definition. 
    %\newpage

    \subsection{Determination of the Mixture Weight} \label{sub:mixture_weight}
        The mixture weight $f_q(w_{i,0})$ as used in Eq.~\ref{eq:general_model} is characterized as the fraction of stars belonging to the mixture component $q$ over an increment $[w_{i,0}, w_{i,0} + \dif{w_{i,0}}]$. Thus, the mixture weight encodes information about the linear density of the stream, which has been shown to depend on substructure in the galaxy (CITE-x). The dependence of the mixture weight on the linear density is captured by the expression
        \begin{equation}\label{eq:fraction_param}
            f_q\left(w_{i,0}\right) = \frac{N_q\left(\phi_1\right)}{N_{\rm tot}\left(\phi_1\right)},
        \end{equation}
        where $N_q\left(\phi_1\right)$ is the number of stars in component $q$ within an interval $[\phi_1, \phi_1 + \dif{\phi_1}]$, and 
        $N_{\rm tot}$ is the total number of stars in the same band (including those in component $q$). The linear  density for component $q$ is then entirely specified by the function $N_q(\phi_1)$.

        The mixture model specified in \S-X only models the weight parameter $f_q$, and not the components of the ratio in \autoref{eq:fraction_param}. However, the linear density $N_q(\phi_1)$ can be readily obtained from $f_q$, provided that $N_{\rm tot}\left(\phi_1\right)$ can be calculated. Importantly, this function does not depend on the parameters of the on-stream or off-stream model. Instead, $N_{\rm tot}(\phi_1)$ is simply the number of stars in a small $\phi_1$ increment. We therefore model $N_{\rm tot}\left(\phi_1\right)$ as a post-processing step using a single normalzing flow in $\phi_1$.

        In particular, for a given field the distribution of $\phi_1$ coordinates is specified by the set $\{\phi_{1,i}\}_i$. A normalizng flow is trained to this dataset, and will satisfy $\int d\phi_1 P_{\phi_i}(\phi_1) = 1$ by construction. In order to re-weight the normalizing flow to obtain $N_{\rm tot}(\phi_1)$, we simply rescale $P_{\phi}(\phi_1)$ by $N_{\rm tot,0}$, which is the total number of stars in the field considered. We can then obtain the linear stream density with
        \begin{equation}
            N_q\left(\phi_1\right) = f_q N_{\rm tot} \left(\phi_1\right)  = f_q \left[N_{\rm tot,0} P_{\phi_1}\left(\phi_1\right) \right] .
        \end{equation}

    \subsection{Astrometric Model} \label{sub:astrometric_model}

        \subsubsection{On-Stream} \label{ssub:astrometric_model_on_stream}

            The astrometric track is modeled with a neural network in each phase space dimension, parameterized as a function of $\phi_1$. The neural network output for a single stream component is the mean location of the stream track in each dimension, $\mbs{\mu}_w(\phi_1 | q_i)$, and its associated covariance matrix $\mbs{\Sigma}_w(\phi_1 | q_i)$. The probability for the data is then
            \begin{equation}
                \pdf^{(S)}(\nth{\mbs{w}} | \mbs{\theta}, \phi_{1,i}, q_i) = \mcal{N}\left(\nth{\mbs{w}} | \mbs{\mu}_w\left(\phi_{1,i} | q_i \right), \mbs{\Sigma}_w\left(\phi_{1,i} | q_i \right) \right),
            \end{equation}
            where $\mbs{\theta}$ is the set of neural network parameters that define $\mbs{\mu}_w$ and $\mbs{\Sigma}_w$.
            In the simplest case, we take $\mbs{\Sigma}_w$ to be diagonal, consisting of both intrinsic dispersion for each dimension of $\mbs{w}$ and the data uncertainties added in quadrature.

            In practice, we work with truncated Gaussians, where the truncation is set by the size of the field in each astrometric dimension. This ensures that the probability density is zero wherever there is no data. Details of the truncated Gaussian are included in \autoref{app:distributions}.

        \subsubsection{Off-stream} \label{ssub:astrometric_model_off_stream}

            We now outline the off-stream astrometric model. We take a similar approach to the on-stream astrometric model discussed in \S\ref{ssub:astrometric_model_on_stream}, though we typically do not characterize the backgrounds with a Gaussian distribution. Instead, we utilize a range of distributions (i.e., exponential, skew-normal) discussed in \autoref{app:distributions}. The parameters of the user-specified background distributions are themselves neural network outputs, such that any $\phi_1$ ``slice" of the background should be characterized by some analytic density, but the evolution of the backgrounds with $\phi_1$ can be complex.

            Let $\mbs{\theta}_{B,j}(\phi_1)$ represent the neural network which outputs a vector of parameters characterizing the background distribution at a given $\phi_1$ for astrometric dimension $j$. The number of outputs for the background model is equal to $\Sigma_{j>0} \mathrm{dim}\left(\mbs{\theta}_{B,j}\right)$ (where $j=0$ corresponds to $\phi_1$, which we do not model). At a given $\phi_1$, the probability density for the astrometric dimension $j>0$ is $\pdf^{(B)}_j\left(\mbs{w}_{n,j} | \mbs{\theta}, \mbs{\theta}_{B,j}, \phi_1 \right)$. We treat each background component as independent, allowing us to write the likelihood as a product over the astrometric dimensions:
            \begin{equation}
                \pdf^{(B)}\left(\mbs{w}_n | \mbs{\theta}, \phi_1 \right) = \prod_{j>0} \pdf_j^{(B)}\left(\mbs{w}_{n,j} | \mbs{\theta}, \mbs{\theta}_{B,j}, \phi_1 \right).
            \end{equation}

    %\newpage

    \subsection{Photometric Model} \label{sub:photometric_model}

        \subsubsection{On-Stream} \label{sub:photometric_model_on_stream}
            In photometric coordinates we model the stream as a single-population isochrone with the possibility of a non-zero distance gradient. We accomplish this by modeling the distance of the isochrone as a neural network output, parametrized as a function of $\phi_1$. The result is a distance track along the stream, estimated from the variation in its isochrone with $\phi_1$.

            At any given $\phi_1$, the isochrone is modeled in absolute magnitudes
            and parameterized by the unit-normalized stellar mass, labeled $\gamma$. For example, increasing $\gamma$ corresponds to movement up the main sequence towards the red giant branch. We denote the isochrone as $\mcal{I(\gamma)}$. The intrinsic dipsersion around the isochrone is $\Sigma_\mcal{I}(\gamma)$. Since the isochrone is in absolute magnitudes, but the data are in apparent magnitudes,
            it is necessary to shift the isochrone by a distance modulus to the predicted distance of the stream, labeled  $\mu(\nth{\phi_1})$.

            For the $n$-th star, the photometric model is:

            \begin{multline}
                P^{(S)}(\mbs{m_i}, \gamma; \mbs{\theta}(\nth{\phi_1}), \nth{\phi_1}) 
                \\ \sim \mcal{N}(\nth{\mbs{m}}; \{\mcal{I(\gamma)} + \mu(\nth{\phi_1}), \sqrt{\Sigma_{\mcal{I}}^2(\gamma) + \Sigma_\mu^2(\nth{\phi_1}) + \Sigma_i^2} \}) \\ \times\pdf(\gamma, \nth{\phi_1}),
            \end{multline}

            where $\pdf(\gamma)$ encodes both the present-day mass function of the stream, as well as the observational constraints (e.g. $g < 22$ mag).

>>>>

\documentclass[twocolumn]{aastex631}
\usepackage{showyourwork}
\usepackage{amsmath, amssymb}
\usepackage{gensymb}
\usepackage{xcolor}
\usepackage{booktabs}
\usepackage{dsfont}
\usepackage[inline]{enumitem}
\usepackage{cleveref}
\usepackage{transparent}
\usepackage{array}
\usepackage{float}

% Change the section names
\renewcommand{\sectionautorefname}{\S\!}%
\renewcommand{\subsectionautorefname}{\S\!\!}%
\renewcommand{\subsubsectionautorefname}{\S\!}%
\renewcommand{\equationautorefname}{Eq.}
\renewcommand{\tableautorefname}{Table}
\renewcommand{\figureautorefname}{Fig.}

\newcommand{\code}[1]{\textsc{#1}}
\newcommand{\package}[1]{\code{#1}}
\newcommand{\stream}[1]{#1}
\newcommand{\dataarchive}[1]{\textit{#1}}
\newcommand{\Gaia}{\dataarchive{Gaia}}
\newcommand{\PanStarrs}{\dataarchive{Pan-STARRS1}}

% Math fonts
\newcommand{\mrm}[1]{\mathrm{#1}}
\newcommand{\mbs}[1]{\boldsymbol{#1}}
\newcommand{\mbf}[1]{\mathbf{#1}}
\newcommand{\mbb}[1]{\mathbb{#1}}
\newcommand{\mfk}[1]{\mathfrak{#1}}
\newcommand{\mcal}[1]{\mathcal{#1}}

% Math symbols
\newcommand{\Exp}[1]{e^{#1}}
\newcommand{\Erf}[1]{\mathrm{erf}\left(#1\right)}
\NewDocumentCommand \dif {o m} {
	\IfNoValueTF{#1}
		{ \mathrm{d}{#2} }
		{ \mathrm{d}^{#1}{\!#2} }
}

% Statistics
\newcommand{\pdf}{p}
\newcommand{\cdf}{P}
\newcommand{\likelihood}{\mcal{L}}
\newcommand{\prior}{\mcal{\pi}}
\newcommand{\posterior}{p}

% Labels
\newcommand{\nth}[1]{{#1}_{\mrm{n}}}  % measurement
\newcommand{\fth}[1]{{#1}_{\mrm{f}}}  % feature
\newcommand{\qth}[1]{{#1}_{\mrm{q}}}  % model index
\newcommand{\unit}[1]{[\text{#1}]}

\newcommand{\smallcomponent}[2]{#2^{\scriptscriptstyle (#1)}}
\newcommand{\cmp}[2]{\smallcomponent{#1}{#2}}
\newcommand{\Scmp}[1]{\cmp{S}{#1}}
\newcommand{\Bcmp}[1]{\cmp{B}{#1}}
\newcommand{\astroM}[1]{{#1}_w}
\newcommand{\photoM}[1]{\smallcomponent{m}{#1}}
\newcommand{\chemoM}[1]{\smallcomponent{z}{#1}}

\newcommand{\Spdf}{\Scmp{\pdf}}
\newcommand{\Bpdf}{\Bcmp{\pdf}}

\newcommand{\parallax}{\varpi}
\newcommand{\sigobs}{{\sigma_*}}

\defcitealias{Ibata+2021}{I21}

@@

% Title
\title{Stream Members Only:\\ Data-Driven Characterization of Stellar Streams with Mixture Density Networks}

% Author list
\author[0000-0003-3954-3291]{Nathaniel Starkman}
\altaffiliation{Co-first author.}
\affiliation{David A Dunlap Department of Astronomy and Astrophysics, University of Toronto, 50 St. George St, Toronto ON M5S 3H4, Canada}

\author[0000-0001-8042-5794]{Jacob Nibauer}
\altaffiliation{Co-first author.}
\affiliation{Department of Astrophysical Sciences, Princeton University, 4 Ivy Ln, Princeton, NJ 08544, USA}

\author[0000-0001-6855-442X]{Jo Bovy}
\affiliation{David A Dunlap Department of Astronomy and Astrophysics, University of Toronto, 50 St. George St, Toronto ON M5S 3H4, Canada}

\author[0000-0003-3613-0854]{Jeremy J. Webb}
\affiliation{Department of Science, Technology and Society, York University, 4700 Keele St, Toronto ON M3J 1P3, Canada}

\author[0000-0001-6584-6144]{Kiyan Tavangar}
\affiliation{Department of Astronomy, Columbia University, 538 West 120th Street, New York, NY 10027, USA}

\author[0000-0003-0872-7098]{Adrian Price-Whelan}
\affiliation{Center for Computational Astrophysics, Flatiron Institute, 162 Fifth Ave, New York, NY 10010, USA}

\author[0000-0002-7846-9787]{Ana Bonaca}
\affiliation{The Observatories of the Carnegie Institution for Science, 813 Santa Barbara Street, Pasadena, CA 91101, USA}

\correspondingauthor{Nathaniel Starkman}
\email{n.starkman@mail.utoronto.ca}

\shortauthors{Starkman and Nibauer et al.}

\begin{abstract}

    Stellar streams are sensitive probes of the Milky Way's gravitational
    potential. The mean track of a stream constrains global properties of the
    potential, while its fine-grained surface density constrains galactic
    substructure. A precise characterization of streams from potentially noisy
    data marks a crucial step in inferring galactic structure, including the
    dark matter, across orders of magnitude in mass scales. Here we present a
    new method for constructing a smooth probability density model of stellar
    streams using all of the available astrometric and photometric data. To
    characterize a stream's morphology and kinematics, we utilize mixture
    density networks to represent its on-sky track, width, stellar number
    density, and kinematic distribution. We model the photometry for each stream
    as a single-stellar population, with a distance track that is simultaneously
    estimated from the stream's inferred distance modulus (using photometry) and
    parallax distribution (using astrometry). We use normalizing flows to
    characterize the distribution of background stars. We apply the method to
    the stream \stream{GD-1}, and the tidal tails of \stream{Palomar 5}. For
    both streams we obtain a catalog of stellar membership probabilities that
    are made publicly available.  Importantly, our model is capable of handling
    data with incomplete phase-space observations, making our method applicable
    to the growing census of Milky Way stellar streams. When applied to a
    population of streams, the resulting membership probabilities from our model
    form the required input to infer the Milky Way's dark matter distribution
    from the scale of the stellar halo down to subhalos.

\end{abstract}

@@

    Stellar streams are the disrupted remnants of globular clusters and
    satellite galaxies, and their census has grown extensively in the Milky Way
    thanks to all-sky astrometric missions like \Gaia{}
    \citep{GaiaCollaboration+2016, GaiaCollaboration+2023} and low-surface
    brightness photometric surveys (e.g., \citealt{Ibata+1998,
    GrillmairDionatos2006, Shipp+2018, Ibata+2021}).  The formation of streams
    can be traced back to a progenitor (a globular cluster or satellite galaxy)
    whose stars are gradually lost to the tidal field of the host galaxy. The
    stars extend along a series of similar orbits, called tidal tails, which can
    sometimes span several tens of degrees across the sky. The orbital proximity
    of stars belonging to a stream makes them sensitive tracers of the mass
    distribution of galaxies, sourced by both the baryonic and dark matter
    components.

    The average properties of a stream (e.g., the mean position and velocity
    track of the tidal tails) can be used to reconstruct the global dark matter
    distribution of the stellar halo (e.g., \citealt{Johnston+1999, Bonaca+2014,
    Bovy2014, Bovy+2016, Nibauer+2022, Koposov+2023}), while local variations in
    the stream track and surface density encode information about the stream's
    progenitor \citep{Kupper+2012, Sanders+2013, Price-Whelan+2014}, encounters
    with baryonic structures such as the stellar bar (e.g., \citealt{Erkal+2017,
    Pearson+2017}), and non-luminous structures like dark matter subhalos (e.g.,
    \citealt{Ibata+2002, Johnston+2002, Carlberg+2012-74820C, Bovy+2017,
    Bonaca+2019, Hermans+2021}).

    While several methods have been developed to model stream properties as a
    function of global halo characteristics (e.g., the mass distribution,
    flattening; \citealt{Johnston+1999, Binney2008, Koposov+2010,
    SandersBinney2013, Bovy2014, Nibauer+2022}) and local substructures (e.g.,
    the bar's pattern speed and the mass of subhalo perturbers;
    \citealt{Antoja+2014, Hattori+2016, Price-Whelan+2016, Bovy+2017,
    Pearson+2017, Bonaca+2019}), these methods often rely on the existence of a
    homogenized stream-tracer population that characterizes a given stream
    across each of the observable phase-space dimensions.  Additionally, because
    even the fine-grained properties of streams are influenced by the detailed
    mass distribution of its host galaxy, it is important to propagate errors
    both in measurement uncertainty and membership uncertainty (i.e., which
    stars belong to the stream versus the background) when attempting to model a
    stream. Otherwise, a poorly characterized stream could bias constraints on,
    e.g., the total mass of the galaxy, the dark matter halo shape, and the mass
    of subhalo perturbers.  In order to generate more precise constraints on the
    structure of the galaxy, it is therefore imperative that a statistically
    sound and homogenized catalog of stellar streams is produced.

    While streams are well described in action-angle coordinates
    \citep[e.g.][]{Bovy2014} or (restricted) $N$-body simulations
    \citep[e.g.][]{Dehnen+2004}, these methods rely on prior assumptions about
    the galaxy and its underlying gravitational potential.  Therefore, it is
    necessary to devise data-driven methods to characterize stellar streams
    without appealing to strong prior assumptions about the galaxy's structure.

    There are a few previous works which have modeled stellar streams in this
    context. \citet{Patrick+2022} developed the method introduced by
    \citet{Erkal+2017}, which is a data-driven, spline-based method for modeling
    the photometric properties and on-sky positions of previously identified
    streams. Their method works by fitting the stream's stellar surface density
    with a series of splines, capturing both surface brightness variations along
    the stream and changes in the position of the stream track and width. Their
    method also utilizes an isochrone fit to a given stream, from which a
    distance modulus can be derived. While this method is very successful at
    characterizing the photometric properties of a stream, it does not consider
    the kinematic dimensions which have become increasingly well-measured with
    successive data releases from \Gaia{}, as well as targeted observations.
    From a population of RR Lyrae stars, \citet{Price-Whelan+2019} utilized a
    mixture model to fit the globular cluster stream Pal-5 in the space of
    proper motions and heliocentric distances.  The mixture modeling approach
    enables a simultaneous fit to both the stream and background density,
    allowing for a cleaner selection of stream stars. However, the track of the
    stream in on-sky coordinates is not fit. The method identified 27 RR Lyrae
    stars consistent with being members of the stream. While useful for
    determining the average properties of a stream (e.g., its mean distance
    track), a limited population of RR Lyrae stars are not sufficient to capture
    the small-scale properties of streams expected to inform constraints on,
    e.g., a population of perturbers in the galaxy.

    Other methods for extracting stellar streams from the data include
    matched-filter algorithms, where over-densities in color-magnitude space are
    inspected as possible clusters undergoing tidal stripping
    \citep{Grillmair+1995, GrillmairJohnson2006, Shipp+2018}.  While the
    matched-filter technique has been extremely successful in stream discovery,
    it is not equipped to characterize streams at the level of detail necessary
    to inform precise constraints on the potential of the galaxy and its
    substructure.

    In order to address the need for a flexible model of stellar streams both in
    photometric and kinematic spaces, in this work we develop a new method for
    characterizing streams and quantifying the membership probability of
    possible stream stars. Our approach provides several key advantage over
    previous stream modeling efforts, specifically in its ability to jointly
    model the kinematic and photometric properties of streams simultaneously
    using a flexible model.  Furthermore, our method is well-suited to fitting
    streams with only partial or incomplete phase-space or photometric
    observations, enabling a full characterization of streams insofar as the
    available data allows.

    \paragraph{Paper organization}

        The paper is organized as follows.  In \autoref{sec:method} we introduce
        our method for characterizing stream tracks and computing stellar
        membership probabilities.  We discuss initial data processing in
        \autoref{sub:method:framing_the_data} and build a probabilistic
        framework to model the stream and background in
        \autoref{sub:method:likelihood_setup}-\autoref{sub:methods:priors}.  In
        \autoref{sec:results_mock} we apply our method to simulated data to
        demonstrate the model with a known ground truth dataset, then to the
        stellar streams \stream{GD-1} (\autoref{sec:results_gd1}) and
        \stream{Pal\,5} (\autoref{sec:results_pal5}).  We compare our model to
        other methods in \autoref{sec:discussion}.  We conclude in
        \autoref{sec:conclusions}, discussing the results of this work and
        future directions.

@@

\newpage
\section{Method} \label{sec:method}

    Below, we offer a concise overview of our approach, followed by a more
    in-depth technical explanation in subsequent sections.  We start, in
    \autoref{sub:method:framing_the_data}, by transforming the astrometric data
    into a stream-oriented on-sky reference frame ($\phi_1, \phi_2, \parallax,
    \mu_{\phi_1}^*, \mu_{\phi_2}, v_r)$, for each stream.  In this frame, the
    stream track lies along $\phi_1$, and the stream density may be described by
    distributions in the other parameters as injective functions of $\phi_1$
    (called conditional distributions).  Likewise, the background of the
    Galactic field may be described with $\phi_1$-conditioned distributions.  We
    will also model the stream in photometric coordinates.  Additional features,
    e.g.  metallicity, can also be modeled along the stream and background as
    functions of $\phi_1$, though we leave this to a future work.

    The distributions for the stream and background are collected in a single
    mixture model that describes the entire field.  We take the parameters of
    the mixture model --- the mixture weights and component distribution
    parameters --- to be general (continuous) functions of $\phi_1$, the
    conditioning variable. This is done using a feed-forward neural network that
    takes $\phi_1$, the conditioning variable, as its input and outputs the
    parameter value(s) that characterize the distribution of stars at a $\phi_1$
    ``slice".  Mixture models of conditional probability distributions using
    neural networks are called Mixture Density Networks (MDNs)
    \citep{Bishop+1994}.  Since the model coefficients may evolve with $\phi_1$,
    MNDs are well suited to characterize streams (and the background) with all
    their variations in position, width, and linear density.

    % sub: framing_the_data (end)

    \subsection{Framing the Data} \label{sub:method:framing_the_data}

        As a stream progenitor orbits the Galactic center of mass, its path can
        be approximated locally as an elliptical segment. The associated arms of
        the stream generally align closely with its orbital plane.  As viewed
        from a Galactocentric reference frame, the stream approximately assumes
        the form of a great circle enabling a 1:1 mapping from some phase
        parameter to a unique position along the stream. Thus in a `stream'
        oriented frame the rotated longitudinal and latitudinal coordinates
        ($\phi_1, \phi_2$) align with the stream such that the stream's phase is
        $\phi_1$ and $\phi_2 \simeq 0$.

        While it would be ideal to characterize streams in this Galactocentric
        frame, transformation to the frame requires knowledge of the distance to
        the Galactic center  \citep{GRAVITY2018, BennettBovy2019, Leung+2022},
        the solar motion around the Galactic center, and accurate distances to
        field stars (among other uncertain quantities). In a heliocentric frame,
        e.g. the standard ICRS\footnote{International Celestial Reference
        System} \citep{ICRS1997}, projection effects can obscure the shape of a
        stream leading to a non-trivial stream track. Therefore in a
        heliocentric frame, there is no assurance of effecting a rotation like
        in the Galactocentric frame that would similarly align the stream in an
        observer's coordinates such that $\phi_2(\phi_1) \approx 0$.
        Thankfully, most streams are adequately distant from both the Galactic
        center and the observer, or exhibit orientations such that a rotation a)
        can be established or b) holds valid. For most of the best-studied
        globular cluster streams -- \stream{GD-1}, \stream{Palomar 5},
        \stream{Jhelum}, etc. -- this holds true. However for some streams of
        dwarf galaxies, like the Magellanic and Sagittarius streams which each
        wrap around the Milky Way \citep{Newberg+2002, Majewski+2003,
        WannierWrixon1972}, no such transformation is possible. More generally,
        a stream's ``sky"-projected path may be kinked, e.g. by subhalo
        interactions, such that it is a many-to-one function in $\phi_1$.

        In this work we consider the vast majority of streams that can be
        described as an injective function, $\phi_2(\phi_1)$.  Since we are
        examining known streams this transformation is known \textit{a priori},
        or may be constructed during pre-processing
        \citep[see][\S2.1]{Starkman+2023}.  We note that, with sufficient prior
        knowledge, many-to-one functions may be broken into injective segments
        and their connection constrained by a prior. Thus the methods developed
        here may be extended and applied even to wrapped and kinked streams.
        However, we leave extending the model framework to a future work.

    \subsection{Likelihood Setup}\label{sub:method:likelihood_setup}

        \begin{figure*}
            \script{pgm.py}
            \centering
            \includegraphics[scale=0.6]{figures/pgm.pdf}
            \caption{%
                Probabilistic Graphical Models (PGMs) of the Mixture Density
                Networks used in this work.
                %
                \textbf{Left}: A PGM of the stream model. The full dataset
                $x_n^{obs}$ is the aggregation of the observed astrometry
                $w_n^{obs}$ and photometry $m_n^{obs}$, along with the
                corresponding observational errors. We model the data as a
                mixture of Gaussians (in astrometry + distance modulus), with
                the Gaussians indexed by $q$. The weight $\alpha$ determines the
                weighted contribution of each Gaussian to each datum. All models
                are conditioned on $\phi_1$.
                %
                \textbf{Right}: a PGM of the background model. The data are
                identical to that of the stream, however the model is not a
                mixture of Gaussians, but of a variety of distributions. We
                further distinguish between trainable distribution parameters
                (circles) and fixed distributions (points), like a pre-trained
                normalizing flow background model, explained in further detail
                in \autoref{sub:method:pre-training_distributions}. Alike to the
                stream model, all background models are conditioned on $\phi_1$.
                % 
            }
            \label{fig:PGM}
        \end{figure*}

        Mixture models are a statistical method to represent a population as a
        set of a sub-populations specified by probability distributions.  These
        distributions are characterized by their parameters and the mixture by a
        corresponding set of mixture coefficients. Scalar mixture models may be
        extended to model completely general conditional distributions by making
        the distribution parameters and mixture coefficients into functions of a
        conditional input \citep{McLachlanBasford1989}.  Mixture Density
        Networks (MDNs) are a machine learning method to implement these general
        mixture models, taking the parameters and coefficients to be the outputs
        of feed-forward neural networks and as such, general continuous
        functions \citep{Bishop1994}.  Functional parameters allow the model to
        capture variation over those parameters that is not possible with
        scalar-valued models. Moreover, by using neural networks the inferred
        functional forms are ``model-free" and driven entirely by the data.

        For the MDN we take $x_{\cdot, 0}$, the first feature column of the data
        $\{\mbf{x}\}$, as the `independent' coordinate over which the MDN
        parameters are conditioned. In the context of streams $x_{\cdot, 0}
        \equiv \phi_1$ is the longitude coordinate in a reference frame rotated
        such that the stream is aligned along $\phi_1$. The parameters of the
        stream, the background field, and the mixing coefficients of the two are
        now general data-driven continuous functions along $\phi_1$.

        The PDF of a general MDN is
        \begin{equation} \label{eq:general_mixture_network}
            \!\!\! \pdf(\nth{\mbs{x}} | \mbs{\theta}(\phi_1))
            \!=\! \sum_{\mrm{q} \in \mbs{I_Q}} \! \qth{f}(\phi_1) \cmp{q}{\pdf}\left(\nth{\mbs{x}}|\qth{\mbs{\theta}}(\phi_1)\right),
        \end{equation}
        where $\cmp{q}{\pdf}$ is the PDF of the $q$-th model and $f_q$ its
        mixing coefficient for all $q$ models in the set of models indexed by
        $\mbs{I_Q}$. $\mbs{\theta}$ is the $\phi_1$-conditional output of the
        feed-forward neural network, and $\nth{\mbs{x}}$ the data on the $n$-th
        star in $\{\mbs{x}\}$, not including $\phi_1$. For convenience we drop
        the implied subscript $n$ in $\phi_1$, the explicit notation of
        $\mbs{\theta}$'s dependence on $\phi_1$, and the index set $\mbs{I}_X$
        in summations over its elements $x$.

        The mixture coefficients $\qth{f}$ are normalized such that
        \begin{equation} \label{eq:mixture_weight_sum}
            \sum_{\mrm{q}} \qth{f}(\phi_1) = 1, \qquad \forall \phi_1
        \end{equation}
        In practice this is enforced by defining a background
        \begin{equation} \label{eq:mixture_weight_normalization}
            f_{b}(\phi_1) = 1 - \sum_{\rm{q}\in \mbs{I_Q}\backslash \{b\}} \qth{f}(\phi_1),
        \end{equation}
        the summation over all models except for some ``background" index $b \in
        \mbs{I_Q}$.  In the context of Galactic streams the ``foreground" is the
        stream itself, while the background is the Galaxy --- bulge, bar, disc,
        halo --- and additional structures, like globular clusters and dwarf
        galaxies, that are not associated with the stream itself.  We discuss
        the mixture weights further in \autoref{sub:method:mixture_weight}.

        In this work we consider primarily astrometric and photometric data,
        (notated $\{\mbs{w}\}$ and $\{\mbs{m}\}$, respectively), but note that
        the MDN framework may be extended to arbitrary feature dimensions.
        Assuming the conditional independence of the astrometry and
        photometry\footnote{ For \Gaia{} data this is true in practice but not
        in detail: there exist small correlations between the astrometric and
        photometric data.}, we split each model into linearly separable
        astrometric and photometric models $\cmp{q}{\pdf} = \cmp{q,w}{\pdf}
        \cmp{q,m}{\pdf}$.

        The likelihood for a single star then becomes, simplifying from
        \autoref{eq:general_mixture_network},
        \begin{equation} \label{eq:general_model}
            \!\!\!\!\! \pdf(\nth{\mbs{w}}, \! \nth{\mbs{m}} | \mbs{\theta})
                \!=\! \sum_{q} \! f_q(\phi_1) \cmp{q,w}{\pdf}(\nth{\mbs{w}}|\mbs{\theta}) \cmp{q,m}{\pdf}(\nth{\mbs{m}}|\mbs{\theta})
        \end{equation}

        Assuming the measurement of each star is independent, the total
        log-likelihood is a sum over all stars:
        \begin{equation} \label{eq:general_likelihood}
            \ln\mcal{L}\left(\{\nth{\mbs{w}},\nth{\mbs{m}}\} | \mbs{\theta}\right) = \sum_n \ln \pdf(\nth{\mbs{w}}, \nth{\mbs{m}} | \mbs{\theta}).
        \end{equation}
